---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)

library(tidyverse)

ggplot2::theme_set(ggprism::theme_prism(palette = "viridis"))

```

# GLabR

<!-- badges: start -->
<!-- badges: end -->

The goal of GLabR is to provide a centralized locations to hold functions that are routinely useful in the Gonzalez Lab at UCSD

## Installation

You can install the development version of GLabR like so:

```{r,eval=FALSE}
devtools::install_github("baynec2/GlabR")
```

## Examples

### Data normalization (batch correction)

The first use case of GlabR is to normalize data that we export from proteome discoverer. Here, a text file containing PSMs is exported and subsequently needs to be processed. To accomplish this, there are a few different steps that need to be followed . These are described below. 

1. We need to combine PSMs from each of the different fractions (there are multiple fractions corresponding to a single sample). This is accomplished using the combine_psm_fractions() function. Essentially, this function filters out PSMs based on certain criteria and then sums the intensities for each protein that they map to.

2. Then we need to normalize our data to account for batch corrections. This is accomplished using the normalize_to_bridge() function. 

  * In some cases, there may not be a bridge channel included, and in these cases we will need to use the normalize_1plex() function instead

3. After this, we still might want to normalize the data further. To get data that is more normally distributed, we can use Leigh-ana's method of box cox normalization through the la_box_cox_norm() function. 


Here we can see an example of the overall workflow using a single plex example set

```{r example}
library(GLabR)

data = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  normalize_1plex()
 
head(data)                   
```


The final norm column has the completely normalized data. 



Here we can see an example of the overall workflow using an example set with multiple plexes 

```{r}
data = read_delim("tests/testdata/normalize_to_bridge/PSM_output.txt") %>% 
  combine_psm_fractions() %>% 
  normalize_to_bridge(bridge_channel_plex = 126)


head(data)
```



If you prefer data in a wide format ( as reported using the previous script), you can specify data_format = "wide" within the normalize to bridge function. This will give a column containing the final_norm values for each Sample/TMT combination. Note that this is also an option for the normalize_1plex function.  An example is shown below

```{r}
data = read_delim("tests/testdata/normalize_to_bridge/PSM_output.txt") %>% 
  combine_psm_fractions() %>% 
  normalize_to_bridge(bridge_channel_plex = 126,data_format = "wide")


head(data)

```



If we wanted to then use the box cox norm method, we could use la_box_cox_norm() function. 

```{r}
data = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  normalize_to_bridge(bridge_channel_plex = 126) %>% 
  la_box_cox_norm()


head(data)
```





Note that when using this function, it is expecting your data to be in the long format. If not, it will not work. If you wish for the results reported from la_box_cox_nrom to be in the wide format, you can specify that like below.

```{r}
data = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  normalize_to_bridge(bridge_channel_plex = 126) %>% 
  la_box_cox_norm(data_format = "wide")


head(data)
```

Out of curiosity, what does the normalized data look like when comparing the two methods. 

```{r}

p1 = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  normalize_to_bridge(bridge_channel_plex = 126) %>% 
  la_box_cox_norm() %>% 
  ggplot(aes(box_cox_scaled_values))+
  geom_histogram()+
  ggtitle("box cox scaled values")



p2 = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  normalize_to_bridge(bridge_channel_plex = 126) %>% 
  ggplot(aes(final_norm))+
  geom_histogram()+
  ggtitle("batch corrected")

p3 = ggpubr::ggarrange(p1,p2)

p3
```
Here we can see the overall distribution of the data. Looks like the box cox does a pretty good job of normalizing the data! 

Note that when I was making this package, I discovered a problem in the script that was originally used to normalize the data. Essentially, the original script was only assigning the NAs from some columns as 1, ultimately causing results to deviate from what was intended (NA from all columns having 1)

Here I have kept the function that produces the same results as the output from the original script (nonnormalizedall.txt) as combine_psm_fractions_replica() for posterity. 


Here we can see the slight differences produced from these different functions. 
```{r}
#using function to replicate new process
new = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  mutate(method = "new")

#using function to replicate old process
old = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions_replica() %>% 
  mutate(method = "old")

#Combining data
comb = bind_rows(new,old) %>% 
  pivot_wider(names_from = method, values_from = value)


#Plotting
p1 = comb %>% 
  ggplot(aes(new,old))+
  geom_point()+
  geom_smooth(method = "lm")

p1
```

As we can see above, while this data is correlated, it is not exactly the same. **Future work should only use the combine_psm_fractions() function.** 

### Protein Idenfification. 

The next step in the process is figuring out what these proteins actually are/ what they do. In order to do this, we can use Uniprot's API to return results pertaining to the proteins. There is a package called UniprotR that handles this, but it was way too slow for long lists of protein IDs. As such, I developed a better solution in the form of the annotate_proteins function. We can annotate these proteins as follows.  



```{r}
protein_list = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  pull(ProteinID) %>% 
  unique()

annotated_proteins = annotate_proteins(protein_list)

head(annotated_proteins)
```
Let's say you wanted to get different information about these proteins. To do that, you can specify what columns to return through the columns argument. This argument takes a string of the column names that you want to add separated by columns. The complete list of field names that are accepted can be found here: https://www.uniprot.org/help/return_fields. 

In this example, lets say we want to get the GO terms for our proteins. We can do that as follows. 

```{r}
protein_list = read_delim("tests/testdata/combine_psm_fractions/PCB002_PSMs_Proteodiscover_output.txt") %>% 
  combine_psm_fractions() %>% 
  pull(ProteinID) %>% 
  unique()

annotated_proteins = annotate_proteins(protein_list,columns ="accession,go_id,go,go_p,go_f,go_c")

head(annotated_proteins)


```




### Phosphoproteomics


The above code is great for looking at proteomics, but sometimes we will want to look at phospo post translational modifications. 

The calc_phospho_ratio function allows us to do this! We can see an example below:


```{r}
proteomics_data = readr::read_delim("tests/testdata/psm_phospho_mod/PCB002_PSMs.txt")
phospho_data = readr::read_delim("tests/testdata/psm_phospho_mod/PCB001_PSM.txt")
metadata = readxl::read_excel("tests/testdata/psm_phospho_mod/metadata.xlsx")
#In this dataset, patient ID is the variable that we want to use to  combine the phospho and proteomic data. 
col_identifying_match = "PatientID"

phospho_ratio = calc_phospho_ratio(proteomics_data,phospho_data,metadata,col_identifying_match,1)

head(phospho_ratio)
```


In short, this code pairs up corresponding proteomics and phosphoproteomics data sets and then calculates the ratio of phosphorylation to using the box cox normalized data.




